Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-12-06 paper/manuscript.tex
updating figures

\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy (FE) calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqeuous to apolar phase~\cite{zwanzig1954hightemperature}. 
The calculations use non-physical intermediate states where atoms are changed from one to another or can be removed or inserted by decoupling or coupling with their environment, Figure~\ref{fig:}. These so-called alchemical states permit converged calculations for large changes, such as the entire removal of a ligand from a binding site, known as an absolute FE calculation, or modifying a substituent between two analogous ligands in a binding site, a relative FE calculation. Two common approaches are free energy perturbation (FEP) and thermodynamic integration (TI) for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, biosimspace}. This foundation work, combined with the methodological, technological and hardware improvements of the last 5-10 years have led to an explosion of interest.  The domain of applicability for these calculations now involves such varied applications as the computation of protein-ligand binding free energies (both relative~\cite{relative-binding-free-energies} and absolute~\cite{absolute-binding-free-energies}), ligand selectivities~\cite{selectivity}, partition and distribution coefficients between different liquid phases (such as octanol-water partition coefficients~\cite{octanol-water-partition}), affinity changes due to mutations in therapeutic targets~\cite{hauser2018predicting,aldeghi2018accurate}, and changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}. As the field of molecular simulation moves towards millisecond timescales accurate FE calculations on even more challenging problems will be in scope. In the meantime, today's user may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.

\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig1/Fig1.pdf}
    \caption{{\bf What is alchemistry?}
    \todo[inline, color=green!20]{JDC: What is the concept for what this figure is supposed to communicate?}
    }
    \label{fig:illustrative-alchemical-transformation}
\end{figure}

@@

    \includegraphics[width=0.95\columnwidth]{paper/figures/fig2/Fig2.pdf}
    \caption{{\bf Illustration of $\lambda$}}
    \label{fig:what_is_lambda}

@@

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{paper/figures/fig3/Fig3.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
    \label{fig:binding-thermodynamic-cycle}
\end{figure}
\todo[inline, color=green!20]{This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

%%%%%%%%%%%%
% Step 0   %
%%%%%%%%%%%%
\section{What can be expected from alchemical simulations? -- Step 0}
\label{sec:step0}

@@

\todo[inline]{JDC: Do we need explicit "Step 0", "Step 1", or can we use some other scheme to clearly delineate the steps one must follow?}

Step 0 is about initially counting the cost of your
project: Can you even hope to tackle the problem you are thinking
about attempting given available resources and, if successful, will it
be worth it?
\subsection*{How accurate are alchemical free energy calculations?}

@@

\subsection*{Drug discovery}

@@

The concerns mentioned so far, such as understanding if the series of molecules is suitable, the simulation timescales relevant to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. However, without venturing too far into details of system setup that is beyond the scope of this article, there are some additional points to consider. Formal charge, or tautomeric state of the small molecules can change within a series. The pKa can change by several log units with only small structural modifications and this is a tactic often used within a congeneric series to modify drug properties. Check up-front because charge or tautomeric change would require modified approaches. A second concern is to try to capture aspects of the bioassay into the simulations, for instance what pH was used for screening? Does the protein X-ray match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer) (cite Perez-Benito Sci Rep 2018, 8, 4883)? Were certain co-factors or partner proteins required to augment activity? 
%

@@

%
%
%
%%%%%%%%%%%%
% Step 1   %
%%%%%%%%%%%%
%
\section{Simulation prerequisites -- Step 1}
\label{sec:step1}
\todo[inline,color={green!20}]{@everyone: please review}
The alchemical free energy protocols, which are discussed in Section~\ref{sec:step2}, which can be used are tied in with what type of free energy you might want to compute, i.e. a free energy of binding or a free energy of hydration. Depending on the type of simulations different considerations have to be made for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
\subsection*{Free energies of binding}
In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates. However, in practice because simulations of thermodynamic states are of finite duration (typically currently 1-100 ns per state) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules. Host-guest or protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation. It is therefore generally important to carefully select input coordinates to obtain satisfactory results. 
You might want to ask yourself the following questions before diving into your simulation setup. 
\begin{itemize}
    \item Do I have one or multiple good host structures? (e.g. a good resolution X-ray crystal of the protein target)
    \item Should I include burried waters, or other small molecules that can be found in an X-ray structure.
    \item Are my ligands part of a congeneric series? (simple R group substitutions around the same scaffold)
    \item Do I have information on one or all of the ligands binding sites (e.g. a X-ray structure)
\end{itemize}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig2/Fig2.pdf}
    \caption{a) Bad alignment of congeneric series b) good alignment}
    \label{fig:not-sure-what-this-is}
\end{figure}

@@

\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig4/topologies.pdf}
    \caption{a) single topology b) dual topology Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
    \label{fig:topology}
\end{figure} 
The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in which functional groups are present; e.g. consider a transformation of benzene (A) to benzyl alcohol (B)~\ref{fig:topology}.
In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group. Bond parameters as well as partial charges are adjusted accordingly. 
Thus in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in fig.~\ref{fig:topology} (a). 
In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{Shirts2012}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{Mobley:2014:J.Comput.AidedMol.Des.}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:topology} (b) depicts how such a dual topology works.

@@

\paragraph{Atom mapping}
Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
Rigorously, this process essentially comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig5/MCS.pdf}
    \caption{a) restrictive MCSS match b) ring breaking}
    \label{fig:mcss}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig6/Fig3.pdf}
    \caption{Absolute thermodynamic cycle}
    \label{fig:absolute-thermodynamic-cycle}
\end{figure}

@@

@@

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%

(Explain the contributions of the different authors here)

@@

@@

\section*{Funding Information}
%%%%%%%
% Authors should acknowledge funding sources here. Reference specific grants.
%%%%%%%
FMS acknowledges the support of NSF grant CHE-1111111.

>>>>

\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy (FE) calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqeuous to apolar phase~\cite{zwanzig1954hightemperature}. 
The calculations use non-physical intermediate states where atoms are changed from one to another or can be removed or inserted by decoupling or coupling with their environment, Figure~\ref{fig:fig1_what_is_alchemy}. These so-called alchemical states permit converged calculations for large changes, such as the entire removal of a ligand from a binding site, known as an absolute FE calculation, or modifying a substituent between two analogous ligands in a binding site, a relative FE calculation. Two common approaches are free energy perturbation (FEP) and thermodynamic integration (TI) for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, biosimspace}. This foundation work, combined with the methodological, technological and hardware improvements of the last 5-10 years have led to an explosion of interest.  The domain of applicability for these calculations now involves such varied applications as the computation of protein-ligand binding free energies (both relative~\cite{relative-binding-free-energies} and absolute~\cite{absolute-binding-free-energies}), ligand selectivities~\cite{selectivity}, partition and distribution coefficients between different liquid phases (such as octanol-water partition coefficients~\cite{octanol-water-partition}), affinity changes due to mutations in therapeutic targets~\cite{hauser2018predicting,aldeghi2018accurate}, and changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}. As the field of molecular simulation moves towards millisecond timescales accurate FE calculations on even more challenging problems will be in scope. In the meantime, today's user may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.

\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig1_what_is_alchemy/Fig1.pdf}    \caption{Illustration of different types of free energies differences that can be computed using alchemical free energy calculations. A: Change in free energy due to a conformational change. B: Partition coefficient such as $\log P$ or $\log D$, here as an example the partition coefficient between methanol and water of a given molecule. C: Free energies of membrane insertions. D: Simulations of portein or other host molecule simulations. E: Absolute free energies of binding of a small molecule to a host (e.g. protein), F: Relative free energy of binding of one molecule with respec to another, here toluene and benzyl alcohol.
    \label{fig:fig1_what_is_alchemy}
    }
\end{figure}

@@

    \includegraphics[width=0.95\columnwidth]{paper/figures/fig2_what_is_lam/Fig2.pdf}
    \caption{By introducing a switching function that depends on a factor $\lambda$ it is possible to describe alchemical intermediates. In A it is shown that at $\lambda$ 0 the molecule is a fully interacting phenol and at $\lambda=1$ a fully interacting benzene. Intermediates can be modelled with a softcore potentials whose shape is illustrated in C with dependence on $\lambda$ and D with dependence on the potential energy. B shows an illustration of the probability distribution of the potential energies as the switching function takes values of $\lambda=0$ to $\lambda=1$, intermediates are required for a sufficient overlap in potential energies to estimate a free energy difference between $\lambda=0$ and $\lambda=1$. }
    \label{fig:fig2_what_is_lambda}

@@

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{paper/figures/fig3_therm_cycle/Fig3.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
    \label{fig:fig3_binding-thermodynamic-cycle}
\end{figure}
\todo[inline, color=green!20]{This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

%%%%%%%%%%%%
% Step 0   %
%%%%%%%%%%%%
\section{What can be expected from alchemical simulations?}
\label{sec:step0}

@@

When starting an alchemcial free energy project you need to determine if an alchemical free energy calculation is the appropriate strategy by initially counting the cost of your
project: Can you even hope to tackle the problem you are thinking
about attempting given available resources and, if successful, will it
be worth it?
\subsection*{How accurate are alchemical free energy calculations?}

@@

%\subsection*{Drug discovery}
\section{Drug discovery}

@@

The concerns mentioned so far, such as understanding if the series of molecules is suitable, the simulation timescales relevant to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. However, without venturing too far into details of system setup, which is beyond the scope of this article, there are some additional points to consider. Formal charge, or tautomeric state of the small molecules can change within a series. The pKa can change by several log units with only small structural modifications and this is a tactic often used within a congeneric series to modify drug properties. Check up-front because charge or tautomeric change would require modified approaches. A second concern is to try to capture aspects of the bioassay into the simulations, for instance what pH was used for screening? Does the protein X-ray match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer) (cite Perez-Benito Sci Rep 2018, 8, 4883)? Were certain co-factors or partner proteins required to augment activity? 
%

@@

%
%
%
%%%%%%%%%%%%
% Step 1   %
%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:step1}
\todo[inline,color={green!20}]{@everyone: please review}
The alchemical free energy protocols, which are discussed in Section~\ref{sec:step2}, which can be used are tied in with what type of free energy you might want to compute, i.e. a free energy of binding or a free energy of hydration. Depending on the type of simulations different considerations have to be made for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
\subsection*{Free energies of binding}
In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates. However, in practice because simulations of thermodynamic states are of finite duration (typically currently 1-100 ns per state) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules. Host-guest or protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation. It is therefore generally important to carefully select input coordinates to obtain satisfactory results. 
You might want to ask yourself the following questions before diving into your simulation setup. 
\begin{itemize}
    \item Do I have one or multiple good host structures? (e.g. a good resolution X-ray crystal of the protein target)
    \item Should I include burried waters, or other small molecules that can be found in an X-ray structure.
    \item Are my ligands part of a congeneric series? (simple R group substitutions around the same scaffold)
    \item Do I have information on one or all of the ligands binding sites (e.g. a X-ray structure)
\end{itemize}

@@

@@

\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig4_topol/Fig4.pdf}
    \caption{top: single topology bottom: dual topology Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
    \label{fig:fig4_topology}
\end{figure} 
The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in which functional groups are present; e.g. consider a transformation of benzene (A) to benzyl alcohol (B)~\ref{fig:topology}.
In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group. Bond parameters as well as partial charges are adjusted accordingly. 
Thus in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in fig.~\ref{fig:fig4_topology} (a). 
In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{Shirts2012}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{Mobley:2014:J.Comput.AidedMol.Des.}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:topology} (b) depicts how such a dual topology works.

@@

\paragraph{Atom mapping}
Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
Rigorously, this process essentially comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig5_mcs/MCS.pdf}
    \caption{A: restrictive MCSS match B: ring breaking}
    \label{fig:fig5_mcss}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig6_thermo_cyc_abs/fig6_therm_abs.pdf}
    \caption{Absolute thermodynamic cycle}
    \label{fig:fig6_absolute-thermodynamic-cycle}
\end{figure}

@@

\subsubsection*{What different samples schemes exist and which one should I use?}

\begin{figure}
    \centering
    \includegraphics[width=0.95*\columnwidth]{paper/figures/fig7_sampl_scheme/sampling_schemes.pd}
    \caption{\textbf{There are 4 different sampling strategies available and implemntations are availble for different software packages.} A: Multiple replicas in parallel at different lambda states. B: Hamiltonian replica exchange scheme. C: Single replica scheme sampling from all $\lambda$ states. D: Non-equilibrium sampling scheme.}
    \label{fig:fig7_sampling_scheme}
\end{figure}

@@

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
(Explain the contributions of the different authors here)

@@

\textit{JS}: created figures 1, 3-6 and contributed to general formatting discussions.

@@

\section*{Funding Information}
%%%%%%%
% Authors should acknowledge funding sources here. Reference specific grants.
%%%%%%%
FMS acknowledges the support of NSF grant CHE-1111111.
ASJSM and JM acknowledge funding through an EPSRC flagship software grant: EP/P022138/1
